---
title: "Analysis Of Key Global Health Indicators"
subtitle: "Relationship Between Socioeconomic Trends and Global Body Mass Index Rates"
author: "Christian Kennedy"
date: "24/04/24"
format: 
 html:
   theme: journal
   backgroundcolor: beige
   toc: true
execute: 
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

# libraries
library(tidyverse)
library(plotly)
library(ggplot2)
library(maps)
library(dplyr)

# data
unicef_indicator_2_unicef_indicator_2_5_ <- read_csv("unicef_indicator_2 - unicef_indicator_2 (5).csv")
unicef_metadata_11_ <- read_csv("unicef_metadata (11).csv")

# transformations
unicef_metadata_11_$year <- as.numeric(unicef_metadata_11_$year)
data_join <- full_join(unicef_indicator_2_unicef_indicator_2_5_, unicef_metadata_11_, by = c("country" = "country", "time_period" = "year"))
```

## Overview

The overall aim behind this report is to provide a comprehensive picture of the positive correlation between **SOCIO-ECONOMIC DISPARITIES** and global **HEALTH INDICATORS**. 

As an essential metric commonly utilised towards evaluating global health trends, the **Body Mass Index (BMI)** levels within various regions and selected countries will be analysed. In evaluating BMI as a health indicator, it is generally acknowledged that **LOW LEVELS** of under 18.5 care classified as **UNDERWEIGHT**, whilst a rate of over 30 is considered **OBESITY**. Throughout this analysis, examples of both ends of the spectrum, heightened and decreased rates, will be provided within various different areas. Upon the conclusion of the report, a thorough understanding aims to be received, centred around offering confirmation to the general perception that countries of which suffer more severely with either **low or high rates** are more negatively impacted by socioeconomic factors, with this report applying both  **GDP PER CAPITA** and **LIFE EXPECTANCY AT BIRTH** to support this viewpoint.



## WORLD MAP: Global BMI Rates in the Available Countries Between 1987 and 2020

```{r}
#| fig-cap: "This global map provides a detailed observation of the average BMI rates in the available countries between 1987 and 2020. Through providing this early introduction, we can identify the countries where the rates are massively higher than reccomended and considered as obesity, and likewise the areas with worryingly low trends where undernourihsment and poverty are rampant, resulting in signifcantlly underweight populations. Both scales of the spectrum indicate extremely concerning trends."

map_world <- map_data("world")

map_data_join <- full_join(
  unicef_indicator_2_unicef_indicator_2_5_, 
  map_world, 
  by = c("country" = "region"), 
  relationship = "many-to-many"
)

ggplot(map_data_join) +
  aes(x = long, y = lat, group = group, fill = obs_value) +
  geom_polygon() +
  scale_fill_gradient(low = "lightblue", high = "red") +
  labs(title = "World Map of global BMI rates", fill = "BMI Rates") +
  theme(text = element_text(family = "serif")) +
  theme_bw()
```
Through a direct focus on this world map, a **comparative analysis** can be provided between the particular countries suffering from both high and low rates of BMI. For instance, it is clear that all of **Argentina, Mongolia and Tunisia** have heightened rates, whilst underdeveloped areas including **Guinea and Niger** record signifcantly lower rates. Given these areas are generally considered poorer, it provides **early evidence** of the extent to the **socio-economic** impact on these findings.


## Time Series Chart: Evolution of BMI Trends in Selected Countries Between 1987 and 2020

```{r}
#| fig-cap: "The time series chart provides a more focused depiction of the overall evolution of global BMI rates between 1987 and 2020. As a means of conducting a regional evaluation, attention is placed on the selection of three countries from each continent, with the overall aim of portraying the fluctuating nature of BMI findings as a global issue ."

countries_of_interest <- c ("Algeria", "Gambia", "Ghana", "Bulgaria", "Georgia", "Germany", "Turkey", "Argentina", "Brazil", "Ecuador", "Bangladesh", "China", "Indonesia","Sri Lanka", "Australia", "Tonga")

timeseries_plot_1 <- data_join %>%
  filter(sex == "Total") %>%
  filter(country %in% countries_of_interest) %>%
  
  ggplot() +
  aes(x = time_period, y = obs_value, group = country, color = country) +
  geom_line() +
  labs(
    x = "Year",
    y = "BMI Rates",
    title = "Evolution of 'Body Mass Index' (BMI) rates by year"
  ) +
  theme_dark() +
  theme(text = element_text(family = "serif"))  

ggplotly(timeseries_plot_1)
```
The most notable component of this chart is the ultimate **positive correlation** between the time period in which both **significant spikes and decreases within this data occur**. Whilst **Europe** acts as an **outlier** due to the limited data, the **global pattern** of BMI trends are **confirmed**. Through evaluating the evolution of each country within the various regions, **fluctuating rates**, with the remainder of this analysis set out ot identify the various factors behind this

## Bar Chart: Comparison Between Males and Females in the Five Most Severely Impacted Countries in 2018

```{r}
#| fig-cap: " This bar chart enables direct focus to be placed on one of the most prominent years for heigtened BMI rates, 2018.Through identifying the five worst performing countries, a cross gender analysis is conducted between males and females and their overall contribution to the total rates in 2018."

top_countries <- c("Argentina", "Mongolia", "Tunisia", "Guinea", "Niger")

filtered_data <- unicef_indicator_2_unicef_indicator_2_5_%>%
  filter(country %in% top_countries & time_period == 2018)

filtered_data %>%
  group_by(country, sex, time_period) %>% 
  ggplot() +
  aes(x = country, y = obs_value, fill = sex) +
  geom_col(position = "dodge", width = 0.90) +
  facet_wrap (2018) +
  labs(
    x = "Five worst performing countries",
    y = "BMI Rates",
    fill = "Gender",
    title = "BMI rates between males and females in 2018"
  ) +
  theme_bw()+
  theme(
    (text = element_text(family = "serif")),
  ) +
  scale_fill_manual(values = c("pink", "lightblue", "darkseagreen3"))
```
The **standout component** of this section is the **equal spread** of BMI levels amongst both males and females. Whilst the data between each country does portray the fact males do display a **slightlly higher** recording than females, the overall trend is quite **balanced**. In addition, this can be identified in **both** areas suffering from high levels and obesity, and those in which the rates are low and undernourishment is prominent. The next section will highlight how the overall outcome of this involves **reduced life expectancy**, regardless of a country being considered **underdeveloped, developing or developed**, as seen through this chart.




## Scatter Plot: Overall Outcome of BMI Trends on Socio-Economic Factors Between 2013 and 2020

```{r}
#| fig-cap: "The final stage of this report involves analysing the wider impact of the levels of BMI on socioe-conomic factors such as GDP per capita and overall life expectancy. Through concluding on this component, we finalise with a comprehensive depection that confirms the overall title of the positive correlation between socieo-economic factors and global health indicators."

countries_of_interest <- c("Argentina", "Mexico","Peru", "Costa Rica", "El Salvador", "Afghanistan", "Bangladesh", "Iraq", "Cambodia", "Cameroon", "Burundi", "Chad", "Tunisia", "Guinea", "Mongolia", "Niger", "Benin", "Georgia", "Albania")
subset_data <- unicef_metadata_11_ %>% filter(country %in% countries_of_interest)

subset_data_filtered <- subset_data %>%
  filter(year >= 2013 & year<= 2020)

ggplot(data = subset_data_filtered) +
  aes(x = `GDP per capita (constant 2015 US$)`, y = `Life expectancy at birth, total (years)`, color = country) +
  geom_point(alpha = 0.5, size = 3.5) +
  facet_wrap(~ year, nrow = 2) +
  scale_x_continuous(
    limits = c(0, 50000),
    breaks = c(20000, 40000),
    labels = scales::unit_format(unit = "K", scale = 0.01)
  ) +
  labs(
    x = "GDP per Capita",
    y = "Life Expectancy",
    title = "Relationship between GDP per capita and life expectancy"
  ) +
  guides(color = "none", size = "none") +
  theme_dark() +
  theme(text = element_text(family = "serif")) 
```
The trend displayed within the scatter plot offers **confirmation** to the overall global perception that countries **suffering more severely from high and low levels of BMI** face **more serious socio-economic factors**. In the case of this example, this incorporates **high levels of GDP per capita and lower life expectancy**. Whilst countries facing low levels of BMI, such as Guinea and Niger, are faced with extremely **low rates of GDP per capita**, the **decreased life expectancy at birth rates**  can actually also be attributed to areas with heightened levels of BMI, with a **mixture of obesity and malnourishment** contributing to the life expectancy trend displayed in this chart. 


## Conclusion & Reccomendations

Overall, this analysis offers a confirmation to the **positive global correlation** between **socio-economic** factors and the **ultimate health standards** within countries accross the world. Through mapping these trends, various conclusions can be drawn. Whilst the data does suggest an overall **equal spread between males and females**, it has identified that the areas and regions suffering from particularly **low and high levels** of BMI, beyond the reccommended rates, subsequently encounter **higher levels of GDP per capita** and **lower levels of life expectancy**. As a direct result of this data, it not only **confirms the ultimate relationship between health indicators and socio-economic factors**, but raises the need for **URGENT action**. Taking this into consideration, we can conclude on the note over the necessity and *urgent need** for **increased policies, assistance and aid** to these particular areas. In doing so, the global efforts and incentive should be based on  **transforming** the worrisome trends identified in this analysis and creating a world with **healthier living standards and populations**.



